Temporal stability in human interaction networks: sector sizes, topological prominence and time activity

### 2015-10-08 paper.tex
revisado corpo do artigo

\begin{abstract}
In this study we report on stable aspects of human interaction networks,
with benchmarks derived from public email lists.
Activity along time and topology evolution were observed by snapshots in a timeline 
and in various scales.
Notably, the activity in timescales, ranging from seconds to months, is practically the same for all networks.
The most important metrics to the dispersion of participants in the topological measures space 
are centrality measurements (degree, strength and betweenness),
followed by symmetry-related metrics and then clustering coefficient.
The activity of participants followed the expected scale-free trace,
thus yielding three classes of vertices by comparison against the Erd\"os-R\'enyi model,
namely hubs, intermediary and periphery.
The relative size of these three sectors did not vary with time and was essentially the same for all email lists.
Typically, 3-12\% of the vertices are hubs, 15-45\% are intermediary and the remainder are peripheral vertices.
Similar results for the distribution of participants in the three categories and for the relative importance of the topological metrics were obtained for 12 additional networks from Facebook, Twitter and Participabr. Consistent with expectations from the literature, these properties may be general for human interaction networks, which has important implications in establishing a typology based on quantitative criteria.
\end{abstract}

@@

\section{Introduction}\label{sec:into}
Studies on human interaction networks have started long before modern computers, dating back to the nineteenth century, while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno in mid twentieth century~\cite{moreno,newmanBook}. With the increasing availability of data related to human interactions, research about these networks has grown continuously. Contributions can now be found in a variety of fields, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic. One of the approaches from an exact science perspective is to represent interaction networks as complex networks~\cite{barabasiHumanDyn,newmanFriendship}, with which 
several features of human interaction have been revealed. For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes. The dynamics of complex networks representing human interaction has also been addressed~\cite{barabasiEvo,newmanEvolving}, but only to a limited extent, since research is normally focused on a particular metric or task, such as accessibility or community detection~\cite{access,newmanModularity}.

In this paper we analyze the evolution of human interaction networks.
Interaction in email lists was the most convenient for deriving results and for benchmarking,
but networks from Facebook, Twitter and Participabr were also considered.
Using a timeline of activity snapshots with a constant number of contiguous messages,
we found remarkable stability for several of the network properties.
Namely, the activity along different timescales exhibit pronounced patterns,
the most basic topological measures always combine into very characteristic principal components,
and the fractions of participants in each of the hubs, intermediady and periphery sectors
are unshaken.
This is not an intuitive result, given that participants transition in network structure constantly.
Because these properties were shared by networks from various sources,
and are consistent with the complex networks literature,
we advocate that the conclusions might be valid for general classes of interaction networks.
In particular, this allows us to 
bridge the gap between data analysis 
and social sciences in the discussion
of types of networks and participants.
Noteworthy, typologies are the canon of scientific literature for classification of human agents,
with pragmatic standards~\cite{myers} and critical paradigms~\cite{adorno,typCanon}.

Section~\ref{sec:related} describes related work,
while details of the data and methods of analysis are given in Section~\ref{sec:data} and Section~\ref{sec:carac}.
Section~\ref{sec:results} brings the results and discussion, leading to Section~\ref{sec:conc} for conclusions.
Subsidiary data analysis,
including directions for video and sound mappings of network structures,
and numeric results for networks from Twitter, Facebook and Participabr,
are given in the Supporting Information.

@@

Unreciprocated edges often exceed 50\% in the analyzed networks, which matches empirical evidence from the literature~\cite{newmanEvolving} and motivated the inclusion of symmetry metrics in our analysis. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were not considered in our study. Gender related behavior in mobile phone datasets was indeed reported~\cite{barabasiSex}, but this was not considered in the present article because email messages and addresses have no gender related metadata~\cite{gmanePack}.

@@

We analyzed many email lists (and data from Twitter, Facebook and Participabr)
and selected four of them in order to make a thorough analysis,
from which general properties can be inferred. These lists are:
\begin{itemize}
	\item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.},
		with participants from different countries with artistic and technological interests.
		English is the most used language. Abbreviated as LAU from now on.
	\item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.},
		with participants from different countries,
		a more technical and less active version of LAU.
		English is the language used the most. 
		Abbreviated as LAD from now on.
	\item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.},
		with computer programmers from different countries.
		English is the most used language.
		Abbreviated as CPP from now on.
	\item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.},
		a Brazilian digital culture interested email list.
		Portuguese is the most used language,
		although Spanish and English are also incident.
		Abbreviated as MET from now on.
\end{itemize}

The first 20,000 messages of each list were considered, with total timespan, authors, threads and missing messages indicated in Table~\ref{tab:genLists}.
We considered 140 additional email lists to
report on the interdependence between the number of participants and 
the number of discussion threads.
Furthermore, 12 additional networks from Facebook (8), Twitter (2) and Participabr (2)
are scrutinized in the Supporting Information document for better hypothesizing
about the generality of the results.

@@

All data and scripts needed to derive results, figures, tables and this article itself
are publicly available.
Email messages are downloadable from the GMANE public database~\cite{GMANEwikipedia}.
Data annotated from Facebook and Twitter are in a public repository~\cite{fbtwData}.
Data from Participabr was used from the linked data/semantic web RDF triples
reported in~\cite{opa} and available in~\cite{datahub}.
Script are delivered through a public domain Python PyPI package and
an open Git repository~\cite{gmanePack}.
This open approach to both data and scripts
reinforces the scientific aspect of the contribution~\cite{openSci}
and lightens ethical and moral issues of researching 
systems constituted of
human individuals~\cite{anPhy,ccs15}.

@@

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted over time as histograms in the scales of seconds,
minutes, hours, days of the week, days of the month, and months of the year.
Most standard measures of location and dispersion, such as usual mean and
standard deviation, hold little meaning in a compact Riemannian manifold.
Equivalent measures were taken using circular statistics,
in which each measurement $t$ (data point) is represented as a unity complex number,
$z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=t\frac{2\pi}{T}$,
and $T$ is the period in which the counting is repeated.
For example, $\theta=12\frac{2\pi}{24}=\pi$ for a message sent at $t=12h$ and given $T=24h$ for days.
The moments $m_n$, lengths of moments $R_n$, mean angles $\theta_\mu$, and rescaled mean angles $\theta_\mu'$ are defined as:

@@

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted,
both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this paper are directed and weighted, the most informative of trivial possibilities.
We did not investigate directed unweighted, undirected weighted,
and undirected unweighted representations of the interaction networks.

The interaction networks were obtained as follows:
a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B.
The reasoning is: if B wrote a response to a message from A,
he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$.
Edges in both directions are allowed.
Each time an interaction occurs, the value of one is added to the edge weight.
Selfloops were regarded as non-informative and discarded.
Inverting edge direction yields the status network:
B read the message and considered what A wrote worth responding, giving status to A,
thus $B\rightarrow A$.
This paper considers by convention the information network as described above ($A\rightarrow B$) and depicted in Figure~\ref{formationNetwork}.
These human social information interaction networks are reported in the literature
as exhibiting scale-free and small-world properties,
as expected for a number of social networks~\cite{bird,newmanBook}.

@@

\begin{itemize}
	\item Degree     $k_i$: number of edges linked to vertex $i$.
	\item In-degree  $k_i^{in}$: number of edges ending at vertex $i$.
	\item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
	\item Strength $s$: sum of weights of all edges linked to vertex $i$.
	\item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
	\item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
	\item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked, i.e. the standard clustering coefficient metric for undirected graphs.
	\item Betweenness centrality $bt_i$: fraction of geodesics that contain vertex $i$. The betweenness centrality index was calculated for weighted digraphs as specified in~\cite{faster}.
\end{itemize}

@@

\subsection{Erd\"os sectioning}\label{sectioning}
In a scale-free network, the peripheral, intermediary and hubs sectors can be derived from a comparison against an Erd\"os-R\'enyi network with the same number of edges and vertices~\cite{3setores}, as depicted in Figure~\ref{fig:setores}. We shall refer to this procedure as \emph{Erd\"os sectioning}, with the resulting sectors being referred to as \emph{Erd\"os sectors} or \emph{primitive sectors}.

The degree distribution $\widetilde{P}(k)$ of an ideal
scale-free network $\mathcal{N}_f(N,z)$ with $N$ vertices and $z$ edges has less
average degree nodes than the distribution $P(k)$ of an Erd\"os-R\'enyi
network with the same number of vertices and edges. Indeed, we define in this work the intermediary sector of a network to be the set of all the nodes whose degree is less abundant in the real network than on the Erd\"os-R\'enyi model:

@@

If $\mathcal{N}_f(N,z)$ is directed and has no self-loops, the probability
of an edge between two arbitrary vertices is $p_e=\frac{z}{N(N-1)}$.
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertices and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability

@@

The lower degree fat tail corresponds to the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is as follows: vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment and as in the Erd\"os R\'enyi model) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in the real network, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser__}
	\caption{Classification of vertices by comparing degree
		distributions.
		The binomial distribution of the Erd\"os R\'enyi network exhibit more intermediary vertices, while a scale-free network, associated with the power-law distribution,
		has more peripheral and hub vertices.
		The sector borders are defined with respect to the intersections of the distributions.
Characteristic degrees are in the compact intervals: $[0,k_L]$, $(k_L,k_R]$, $(k_R,k_{max}]$ for the periphery,
intermediary and hubs sectors, the ``Erd\"os sectors''.
		The connectivity distribution of empirical interaction networks, e.g. derived from email lists,
		can be sectioned by comparison against the associated binomial distribution with the same number of vertices and edges.
		In this figure, a snapshot of 1000 messages from CPP list yield the degree distribution
		of an interaction network of 98 nodes and 235 edges.
		A throughout exposition of the method is exposed in Section~\ref{sectioning}.}
		\label{fig:setores}
\end{figure}
To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network.
The range $\Delta$ of incident values should be partitioned in $m$ parts $\Delta=\cup_{i=1}^m \Delta_i$,
with $\Delta_i\cap \Delta_j=\emptyset \; \forall\; i \neq j$:
%\begin{equation}
%\begin{split}
\begin{align}
\Delta_i =\Biggl\{ k\;\; & | & \overline{\Delta}_{i-1}< &\, k\leq j \text{ and }\;\;\;\;\;\;\;\;\;\;\;\;\nonumber\\
		      & \Biggl[ & N - \sum_{k=0}^{\overline{\Delta}_{i-1}} & \eta_k< \eta \text{ or }\\
&	\Bigl[ & \sum_{k=\overline{\Delta}_{i-1}+1}^j &  \eta_k \geq \eta \text{ and }\;\;\;\;\;\;\nonumber\\
& \Bigl( &\sum_{k=\overline{\Delta}_{i-1}+1}^{j-1} &  \eta_k < \eta \text{ or }   j=\overline{\Delta}_{i-1}+1 \Bigr) \;\Bigr] \Biggr] \Biggr\}\nonumber
\end{align}
%\end{split}
%\end{equation}

@@

If strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used for comparison, with $\overline{w}=2\frac{z}{\sum_is_i}$ the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees ($k^{in}$, $k^{out}$) comparison of the real network should be made with
\begin{equation}
	\hat{P}(k^{way})=\binom{N-1}{k^{way}}p_e^k(1-p_e)^{N-1-k^{way}},
\end{equation}

@@

Since different metrics can be used to identify the three types of vertices, compound criteria were defined for conveniently analysing 
networks with a low number of messages,
such as in Section~\ref{si:frac} of the Supporting Information.
For example, a very stringent criterion can be used, according to which a vertex is only regarded as pertaining to a sector if it is so for all the metrics. After a careful consideration of possible combinations, these were reduced to six:

\begin{itemize}
	\item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified do not usually reach 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics. Therefore, a vertex may belong to more than one class, and the total number of members may exceed 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
	\item Inclusivist cascade $C_4$: vertices are hubs if they are classified as such according to any of the metrics. The remaining vertices are classified as intermediary if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
	\item Exclusivist externals $C_5$: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes for any metric. The rest of the nodes are classified as intermediary.
	\item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
\end{itemize}

Using equations~\ref{eq:part}, these compound criteria $C_\delta$, with $\delta$ integer in the interval $1\leq\delta\leq6$, can be described as:

@@

Let $\mathbf{X}=\{X[i,j]\}$ be a matrix of all vertices $i$ and the respective
values for each metric $j$, 
$\mu_X [j]=\frac{\sum_i X[i,j]}{I}$ the mean of metric $j$ over all $I$ vertices, 
$\sigma_X [j]=\sqrt{\frac{\sum_i (X[i,j]-\mu_X [j])^2}{I}}$ the standard deviation of metric $j$,
and $\mathbf{X'}=\{X'[i,j]\}=\left\{\frac{X[i,j]-\mu_X[j]}{\sigma_X[j]}\right\}$ 
the matrix with the \emph{z-score} of each metric. 
Let $\mathbf{V}=\{V[j,k]\}$ be the matrix $J\times J$ of eigenvectors
of the covariance matrix $\mathbf{C}$
of $\mathbf{X'}$, one eigenvector per column.
Each eigenvector combines the original measures into one principal component, therefore
$V'[j,k]=100\frac{|V[j,k]|}{\sum_{j'} |V[j',k]|}$
is the percentage of the principal component $k$
that is directly proportional to the measure $j$.
%With $k$ eigenvectors 
%$D[k]$,
%it is enough to 
Be $\mathbf{D}=\{D[k]\}$ the eigenvalues associated to the eigenvectors $\mathbf{V}$,
then $D'[k]=100*\frac{D[k]}{\sum_{k'}D[k']}$
is the percentage of total dispersion of the system that the principal component $k$
is responsible for.
We consider, in general, the three greatest eigenvalues and
the respective eigenvectors in percentages:
$\{(D'[k],\;V'[j,k])\}$.
These usually sum between 60 and 95\% of the dispersion
and reveal patterns for a sound analysis.
In particular, 
given $L$ snapshots $l$ of the interaction network,
we are interested in the mean
$\mu_{V'}[j,k]$
and the standard deviation $\sigma_{V'}[j,k]$ 
of the contribution of metric $j$ to the principal component $k$,
and the mean
$\mu_{D'}[k]$
and the standard deviation 
$\sigma_{D'}[k]$
of the contribution of the component $k$ to the dispersion
of the system:

>>>>

\begin{abstract}
In this study we report on stable aspects of human interaction networks,
with benchmarks derived from public email lists.
Activity along time and topology evolution were observed by snapshots in a timeline 
and in various scales.
Notably, the activity in timescales, ranging from seconds to months, is practically the same for all networks.
The most important metrics to the dispersion of participants in the topological measures space 
are centrality measurements (degree, strength and betweenness),
followed by symmetry-related metrics and then clustering coefficient.
The activity of participants followed the expected scale-free trace,
thus yielding the hubs, intermediary and periphery classes of vertices by comparison against the Erd\"os-R\'enyi model.
The relative sizes of these three sectors were essentially the same for all email lists and the same along time.
Typically, 3-12\% of the vertices are hubs, 15-45\% are intermediary and 44-81\% are peripheral vertices.
Similar results for the distribution of participants in the three categories
and for the relative importance of the topological metrics were obtained for 12 additional networks
from Facebook, Twitter and Participabr.
Consistent with expectations derived from the literature,
these properties may be general for human interaction networks,
which has important implications in establishing a typology of participants based on quantitative criteria.
\end{abstract}

@@

\section{Introduction}\label{sec:into}
First studies explicitly about human interaction networks 
date from the nineteenth century while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno in mid twentieth century~\cite{moreno,newmanBook}.
With the increasing availability of data related to human interactions, research about these networks has grown continuously. Contributions can now be found in a variety of fields, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic. One of the approaches from an exact science perspective is to represent interaction networks as complex networks~\cite{barabasiHumanDyn,newmanFriendship}, with which 
several features of human interaction have been revealed. For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes. The dynamics of complex networks representing human interaction has also been addressed~\cite{barabasiEvo,newmanEvolving}, but only to a limited extent, since research is normally focused on a particular metric or task, such as accessibility or community detection~\cite{access,newmanModularity}.

In this paper we analyze the evolution of human interaction networks.
Interaction in email lists was the most convenient for deriving results and for benchmarking,
but networks from Facebook, Twitter and Participabr were also considered.
Using a timeline of activity snapshots with a constant number of contiguous messages,
we found remarkable stability for several of the network properties.
Namely, the activity along different timescales exhibit pronounced patterns,
the most basic topological measures always combine into very characteristic principal components,
and the fractions of participants in each of the hubs, intermediady and periphery Erd\"os sectors
are unshaken.
This is not an intuitive result, given that participants transition in network structure constantly.
Because these properties were shared by networks from various sources,
and are consistent with the complex networks literature,
we advocate that the conclusions might be valid for general classes of interaction networks.
In particular, this allows us to 
bridge the gap between data analysis 
and social sciences in the discussion
of types of networks and participants.
Noteworthy, typologies are the canon of scientific literature for the classification of human agents,
with pragmatic standards~\cite{myers} and critical paradigms~\cite{adorno,typCanon}.

Section~\ref{sec:related} describes related work,
while data, scripts and methods of analysis are given in Section~\ref{sec:data} and Section~\ref{sec:carac}.
Section~\ref{sec:results} reports results and discussion, leading to Section~\ref{sec:conc} for conclusions.
Subsidiary data analysis,
including directions for video and sound mappings of network structures,
and numeric results for networks from Twitter, Facebook and Participabr,
are given in the Supporting Information document.

@@

Unreciprocated edges often exceed 50\% in the analyzed networks, which matches empirical evidence from the literature~\cite{newmanEvolving} and motivated the inclusion of symmetry metrics in our analysis. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were not considered in our study. Gender related behavior in mobile phone datasets was indeed reported~\cite{barabasiSex} but is not relevant for the present article because email messages and addresses have no gender related metadata~\cite{gmanePack}.

@@

We analyzed many email lists (and data from Twitter, Facebook and Participabr)
and selected four of them in order to expose a thorough analysis,
from which general properties can be inferred. These lists are:
\begin{itemize}
	\item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.},
		with participants from different countries with artistic and technological interests.
		English is the most used language. Abbreviated as LAU from now on.
	\item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.},
		with participants from different countries,
		a more technical and less active version of LAU.
		English is the language used the most. 
		Abbreviated as LAD from now on.
	\item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.},
		with computer programmers from different countries.
		English is the most used language.
		Abbreviated as CPP from now on.
	\item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.},
		a Brazilian digital culture interested email list.
		Portuguese is the most used language,
		although Spanish and English are also incident.
		Abbreviated as MET from now on.
\end{itemize}

The first 20,000 messages of each list were considered, with total timespan, authors, threads and missing messages indicated in Table~\ref{tab:genLists}.
We considered 140 additional email lists to
report on the interdependence between the number of participants and 
the number of discussion threads.
Furthermore, 12 networks from Facebook (8), Twitter (2) and Participabr (2)
are scrutinized in the Supporting Information document for better hypothesizing
about the generality of the results.

@@

All data and scripts needed to derive results, figures, tables and this article itself
are publicly available.
Email messages are downloadable from the GMANE public database~\cite{GMANEwikipedia}.
Data annotated from Facebook and Twitter are in a public repository~\cite{fbtwData}.
Data from Participabr was used from the linked data/semantic web RDF triples
reported in~\cite{opa} and available in~\cite{datahub}.
Computer scripts are delivered through a public domain Python PyPI package and
an open Git repository~\cite{gmanePack}.
This open approach to both data and scripts
reinforces the scientific aspect of the contribution~\cite{openSci}
and mitigates ethical and moral issues of researching 
systems constituted of
human individuals~\cite{anPhy,ccs15}.

@@

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted over time as histograms in the scales of seconds,
minutes, hours, days of the week, days of the month, and months of the year.
Most standard measures of location and dispersion, such as usual mean and
standard deviation, hold little meaning in a compact Riemannian manifold.
Equivalent measures were taken using circular statistics,
in which each measurement $t$ is represented as a unity complex number,
$z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=t\frac{2\pi}{T}$,
and $T$ is the period in which the counting is repeated.
For example, $\theta=12\frac{2\pi}{24}=\pi$ for a message sent at $t=12h$ and given $T=24h$ for days.
The moments $m_n$, lengths of moments $R_n$, mean angles $\theta_\mu$, and rescaled mean angles $\theta_\mu'$ are defined as:

@@

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted,
both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this paper are directed and weighted, the most informative of the possibilities.
We did not investigate directed unweighted, undirected weighted,
and undirected unweighted representations of the interaction networks.

The interaction networks were obtained as follows:
a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B.
The reasoning is: if B wrote a response to a message from A,
he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$.
Edges in both directions are allowed.
Each time an interaction occurs, the value of one is added to the edge weight.
Selfloops were regarded as non-informative and discarded.
Inverting edge direction yields the status network:
B read the message and considered what A wrote worth responding, giving status to A,
thus $B\rightarrow A$.
This paper considers by convention the information network as described above ($A\rightarrow B$) and depicted in Figure~\ref{formationNetwork}.
These interaction networks are reported in the literature
as exhibiting scale-free and small-world properties,
as expected for a number of social networks~\cite{bird,newmanBook}.

@@

\begin{itemize}
	\item Degree     $k_i$: number of edges linked to vertex $i$.
	\item In-degree  $k_i^{in}$: number of edges ending at vertex $i$.
	\item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
	\item Strength $s_i$: sum of weights of all edges linked to vertex $i$.
	\item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
	\item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
	\item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked, i.e. the standard clustering coefficient metric for undirected graphs.
	\item Betweenness centrality $bt_i$: fraction of geodesics that contain vertex $i$. The betweenness centrality index was calculated for weighted digraphs as specified in~\cite{faster}.
\end{itemize}

@@

\subsection{Erd\"os sectioning}\label{sectioning}
In a scale-free network, the peripheral, intermediary and hubs sectors can be derived from a comparison against an Erd\"os-R\'enyi network with the same number of edges and vertices~\cite{3setores}, as depicted in Figure~\ref{fig:setores}. We refer to this procedure as \emph{Erd\"os sectioning}, with the resulting sectors regarded as \emph{Erd\"os sectors}.

The degree distribution $\widetilde{P}(k)$ of a
scale-free network $\mathcal{N}_f(N,z)$ with $N$ vertices and $z$ edges has less
average degree nodes than the distribution $P(k)$ of an Erd\"os-R\'enyi
network with the same number of vertices and edges. Indeed, we define in this work the intermediary sector of a network to be the set of all the nodes whose degree is less abundant in the real network than on the Erd\"os-R\'enyi model:

@@

If $\mathcal{N}_f(N,z)$ is directed and has no self-loops, the probability of the existence
of an edge between two arbitrary vertices is $p_e=\frac{z}{N(N-1)}$.
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertices and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability

@@

The lower degree fat tail corresponds to the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$.
The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$.
The reasoning for this classification is as follows: vertices so connected that they are virtually inexistent in networks connected without distinction of the vertices, i.e. without preferential attachment and as in the Erd\"os-R\'enyi model, are correctly associated to the hubs sector.
Vertices with very few connections, which are way more abundant than expected in the Erd\"os-R\'enyi model, are assigned to the periphery.
Vertices with degree values predicted as the most abundant in a Erd\"os-R\'enyi model,
near the average, and less frequent in the real network, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser__}
	\caption{Classification of vertices by comparing degree
		distributions.
		The binomial distribution of the Erd\"os-R\'enyi network model exhibit more intermediary vertices,
		while a scale-free network, associated with the power-law distribution,
		has more peripheral and hub vertices.
		The sector borders are defined with respect to the intersections of the distributions.
Characteristic degrees are in the compact intervals: $[0,k_L]$, $(k_L,k_R]$, $(k_R,k_{max}]$ for the periphery,
intermediary and hubs sectors, the ``Erd\"os sectors''.
		The connectivity distribution of empirical interaction networks, e.g. derived from email lists,
		can be sectioned by comparison against the associated binomial distribution with the same number of vertices and edges.
		In this figure, a snapshot of 1000 messages from CPP list yield the degree distribution
		of an interaction network of 98 nodes and 235 edges.
		A throughout exposition of the method is exposed in Section~\ref{sectioning}.}
		\label{fig:setores}
\end{figure}
To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network.
The range $\Delta$ of incident values should be partitioned in $m$ parts $\Delta=\cup_{i=1}^m \Delta_i$,
with $\Delta_i\cap \Delta_j=\emptyset \; \forall\; i \neq j$ and:
%\begin{equation}
%\begin{split}
\begin{align}
\Delta_i =\Biggl\{ k\;\; & | & \overline{\Delta}_{i-1}< &\, k\leq j \text{ and }\;\;\;\;\;\;\;\;\;\;\;\;\nonumber\\
		      & \Biggl[ & N - \sum_{k=0}^{\overline{\Delta}_{i-1}} & \eta_k< \eta \text{ or }\\
&	\Bigl[ & \sum_{k=\overline{\Delta}_{i-1}+1}^j &  \eta_k \geq \eta \text{ and }\;\;\;\;\;\;\nonumber\\
& \Bigl( &\sum_{k=\overline{\Delta}_{i-1}+1}^{j-1} &  \eta_k < \eta \text{ or }   j=\overline{\Delta}_{i-1}+1 \Bigr) \;\Bigr] \Biggr] \Biggr\}\nonumber
\end{align}
%\end{split}
%\end{equation}

@@

If strength $s$ is used for comparison, $P$ remains the same, but $P(\kappa_i)$ with $\kappa_i=\frac{s_i}{\overline{w}}$ should be used, where $\overline{w}=2\frac{z}{\sum_is_i}$ is the average weight of an edge and $s_i$ the strength of vertex $i$. For in and out degrees ($k^{in}$, $k^{out}$), comparison of the real network should be made against
\begin{equation}
	\hat{P}(k^{way})=\binom{N-1}{k^{way}}p_e^k(1-p_e)^{N-1-k^{way}},
\end{equation}

@@

Since different metrics can be used to identify the three types of vertices,
more than one metric can be used simultaneously, which is convenient when analysing 
small networks,
such as the cases where $ws=50$ in Section~\ref{si:frac} of the Supporting Information.
%For example, a very stringent criterion can be used, according to which a vertex is only regarded as pertaining to a sector if it is so for all the metrics.
After a careful consideration of possible combinations, these were reduced to six:

\begin{itemize}
	\item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics.
		In this case, vertices classified do not usually reach $N$ (or 100\%),
		which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics.
		Therefore, a vertex may belong to more than one class, and the total number of memberhips may exceed $N$ (or 100\%),
		which is indicated by a black line in Figure~\ref{fig:sectIL}.
	\item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics.
		Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics.
		The remaining vertices are regarded as peripheral.
	\item Inclusivist cascade $C_4$: vertices are hubs if they are classified as such according to any of the metrics.
		The remaining vertices are intermediary if they belong to this category for any of the metrics.
		Peripheral vertices are those which are classified as such with respect to all metrics.
	\item Exclusivist externals $C_5$: vertices are hubs if they are classified as such according to all the metrics.
		The remaining vertices are peripheral if they are peripheral or hubs for all metrics.
		The rest of the nodes are intermediary.
	\item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices are peripheral if they are classified as such according to any metric. The rest of the vertices are intermediary.
\end{itemize}

Using equations~\ref{eq:part}, these \emph{compound criteria} $C_\delta$, with $\delta$ integer in the interval $1\leq\delta\leq6$, can be specified as:

@@

Let $\mathbf{X}=\{X[i,j]\}$ be a matrix of all vertices $i$ and the respective
values for each metric $j$, 
$\mu_X [j]=\frac{\sum_i X[i,j]}{I}$ the mean of metric $j$ over all $I$ vertices, 
$\sigma_X [j]=\sqrt{\frac{\sum_i (X[i,j]-\mu_X [j])^2}{I}}$ the standard deviation of metric $j$,
and $\mathbf{X'}=\{X'[i,j]\}=\left\{\frac{X[i,j]-\mu_X[j]}{\sigma_X[j]}\right\}$ 
the matrix with the \emph{z-score} of each metric. 
Let $\mathbf{V}=\{V[j,k]\}$ be the matrix $J\times J$ of eigenvectors
of the covariance matrix $\mathbf{C}$
of $\mathbf{X'}$, one eigenvector per column.
Each eigenvector combines the original measures into one principal component, therefore
$V'[j,k]=100\frac{|V[j,k]|}{\sum_{j'} |V[j',k]|}$
is the percentage of the principal component $k$
that is directly proportional to the measure $j$.
%With $k$ eigenvectors 
%$D[k]$,
%it is enough to 
Be $\mathbf{D}=\{D[k]\}$ the eigenvalues associated to the eigenvectors $\mathbf{V}$,
then $D'[k]=100\frac{D[k]}{\sum_{k'}D[k']}$
is the percentage of total dispersion of the system that the principal component $k$
is responsible for.
We consider, in general, the three greatest eigenvalues and
the respective eigenvectors in percentages:
$\{(D'[k],\;V'[j,k])\}$.
These usually sum between 60 and 95\% of the dispersion
and reveal patterns for a first analysis.
In particular, 
given $L$ snapshots $l$ of the interaction network,
we are interested in the mean
$\mu_{V'}[j,k]$
and the standard deviation $\sigma_{V'}[j,k]$ 
of the contribution of metric $j$ to the principal component $k$,
and the mean
$\mu_{D'}[k]$
and the standard deviation 
$\sigma_{D'}[k]$
of the contribution of the component $k$ to the dispersion
of the system:
